---
title: "Barcelona Cycling Network Connectivity (LCC)"
format: html
---

This document shows how to calculate a **cycling network connectivity indicator** for Barcelona using the **Largest Connected Component (LCC)** of the cycling infrastructure network.

The indicator measures the **share of cycling infrastructure length that belongs to the main connected component** of the network.

# Indicator

Connectivity is defined as:

$$
\text{Connectivity} = \frac{\text{Length of cycling network in the largest connected component}}{\text{Total cycling network length}}
$$

Values closer to **1** indicate a more connected network.

In this document, we first show the method for **one year (2025)**, then calculate the indicator for **census tracts**, and finally repeat the same steps for **all years**, so that we can see the trend.

# Data

Cycling infrastructure data are obtained from [**Open Data BCN**](https://opendata-ajuntament.barcelona.cat/data/es/dataset/carril-bici), which provides shapefiles by semester.

For consistency, we use the fourth-quarter (T4) dataset for each year from **2017 to 2025**.

Census tract boundaries are obtained from the geopackage `Revisió_connectivitat.gpkg`.

# 1. Load packages and data

```{r}
#| message: false
#| warning: false
#| paged-print: false
library(sf)
library(igraph)
library(dplyr)
library(leaflet)
library(ggplot2)

segments <- st_read(
  "data/2025_4T_CARRIL_BICI/2025_4T_CARRIL_BICI.shp",
  quiet = TRUE
)

tracts <- st_read(
  "data/Revisió_connectivitat.gpkg",
  layer = "Seccions_censals",
  quiet = TRUE
)
```

# 2. Prepare the cycling network for 2025

We first transform the cycling network to a projected CRS so that distances and lengths are measured in metres.

```{r}
segments <- segments |>
  st_transform(25831) |>
  st_cast("LINESTRING")
```

# 3. Identify the largest connected component

To identify connected parts of the network, we create a small buffer around each segment and check which segments touch each other.

```{r}

buffer_dist <- 5

buf <- st_buffer(segments, buffer_dist)

touch <- st_intersects(buf)

g <- graph_from_adj_list(touch, mode = "all")

segments$component <- components(g)$membership
segments$len <- as.numeric(st_length(segments))
```

Now we identify which component has the greatest total length.

```{r}
lcc_id <- segments |>
  st_drop_geometry() |>
  summarise(len = sum(len), .by = component) |>
  slice_max(len, n = 1) |>
  pull(component)

segments$LCC <- segments$component == lcc_id
```

# 4. Calculate city-level connectivity for 2025

```{r}
connectivity <- sum(segments$len[segments$LCC]) / sum(segments$len)

connectivity
```

# 5. Visualise the main connected component

In the map below:

- **grey** shows all cycling segments

- **red** shows the segments that belong to the largest connected 
component

```{r}
segments_ll <- st_transform(segments, 4326)

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolylines(data = segments_ll, color = "grey", weight = 3) |>
  addPolylines(data = segments_ll[segments_ll$LCC, ], color = "red", weight = 3)
```

# 6. Calculate connectivity by census tract for 2025

Now we calculate, for each census tract, the share of cycling infrastructure length that belongs to the city's largest connected component.

```{r}
tracts_m <- tracts |>
  st_transform(25831) |>
  select(MUNDISSEC)

segments_tract <- st_intersection(segments, tracts_m)

segments_tract$len <- as.numeric(st_length(segments_tract))

connectivity_tract <- segments_tract |>
  st_drop_geometry() |>
  summarise(
    total_len = sum(len),
    lcc_len = sum(len[LCC]),
    .by = MUNDISSEC
  ) |>
  mutate(
    connectivity = lcc_len / total_len
  )

connectivity_tract
```

# 7. Map connectivity by census tract for 2025

```{r}
tracts_connectivity <- tracts_m |>
  left_join(connectivity_tract, by = "MUNDISSEC")

tracts_ll <- st_transform(tracts_connectivity, 4326)

pal <- colorNumeric(
  palette = "viridis",
  domain = c(0, 1),
  na.color = "transparent"
)

leaflet(data = tracts_ll) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor = ~pal(connectivity),
    fillOpacity = 0.7,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "MUNDISSEC: ", MUNDISSEC,
      "<br>Connectivity: ", round(connectivity, 3)
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~connectivity,
    position = "bottomright",
    title = "Connectivity",
    opacity = 0.7
  )
```

# 8. Repeat the calculation for all years

After understanding the method for one year, we can repeat the same calculation for all years.

```{r}
years <- 2017:2025

results_tract <- vector("list", length(years))
results_city <- vector("list", length(years))

for (i in seq_along(years)) {

  cycle_year <- st_read(
    paste0("data/", years[i], "_4T_CARRIL_BICI/", years[i], "_4T_CARRIL_BICI.shp"),
    quiet = TRUE
  )

  segments <- cycle_year |>
    st_transform(25831) |>
    st_cast("LINESTRING")

  buf <- st_buffer(segments, buffer_dist)
  touch <- st_intersects(buf)
  g <- graph_from_adj_list(touch, mode = "all")

  segments$component <- components(g)$membership
  segments$len <- as.numeric(st_length(segments))

  lcc_id <- segments |>
    st_drop_geometry() |>
    summarise(len = sum(len), .by = component) |>
    slice_max(len, n = 1) |>
    pull(component)

  segments$LCC <- segments$component == lcc_id

  results_city[[i]] <- segments |>
    st_drop_geometry() |>
    summarise(
      total_len = sum(len),
      lcc_len = sum(len[LCC])
    ) |>
    mutate(
      connectivity = lcc_len / total_len,
      year = years[i]
    )

  segments_tract <- st_intersection(segments, tracts_m)
  segments_tract$len <- as.numeric(st_length(segments_tract))

  results_tract[[i]] <- segments_tract |>
    st_drop_geometry() |>
    summarise(
      total_len = sum(len),
      lcc_len = sum(len[LCC]),
      .by = MUNDISSEC
    ) |>
    mutate(
      connectivity = lcc_len / total_len,
      year = years[i]
    )
}

connectivity_all_years <- bind_rows(results_tract)
connectivity_city <- bind_rows(results_city)
```

# 9. Plot city-level connectivity over time

```{r}
ggplot(connectivity_city, aes(year, connectivity)) +
  geom_line() +
  geom_point()
```

