---
title: "Barcelona Cycling Network Connectivity (LCC)"
format: gfm
editor: visual
---

This repository computes a **cycling network connectivity indicator** for Barcelona based on the **Largest Connected Component (LCC)** of the cycling infrastructure network.

The indicator measures the **proportion of the cycling network that belongs to the main connected component**, capturing the degree of **network fragmentation**.

Cycling infrastructure data come from **BCN Open Data**.

# Indicator

Connectivity is defined as:

$$
Connectivity = \frac{\text{Length of cycling network in the largest connected component}}{\text{Total cycling network length}}
$$

Interpretation:

| Value | Meaning |
|:---|:---|
| ~1 | Highly connected network |
| 0.6-0.8 | Moderate fragmentation |
| <0.5 | Highly fragmented network |

This indicator reflects the **continuity of the cycling network**, not just the total amount of infrastructure.

# Data

Cycling infrastructure data are obtained from **BCN Open Data**.

Dataset:
**Carrils bici de Barcelona**

The dataset represents the official cycling infrastructure network maintained by the City of Barcelona.

# Workflow

1. Load cycling infrastructure data
2. Clean and prepare the network
3. Convert the network to a graph
4. Identify connected components
5. Compute total network length
6. Compute length of the largest connected component
7. Calculate the connectivity indicator

```{r}
library(sf)
library(leaflet)

# st_layers("data/Revisió_connectivitat.gpkg")

cycle <- st_read("data/2025_4T_CARRIL_BICI/2025_4T_CARRIL_BICI.shp")

tracts <- st_read(
  "data/Revisió_connectivitat.gpkg",
  layer = "Seccions_censals"
)

# work in projected CRS for metric buffers
cycle_m <- st_transform(cycle, 25831)
tracts_m <- st_transform(tracts, 25831)

# optional buffers
cycle_buf200 <- st_buffer(cycle_m, 200)
tracts_buf50 <- st_buffer(tracts_m, 50)

# back to leaflet CRS
cycle_ll <- st_transform(cycle_m, 4326)
tracts_ll <- st_transform(tracts_m, 4326)
cycle_buf200_ll <- st_transform(cycle_buf200, 4326)
tracts_buf50_ll <- st_transform(tracts_buf50, 4326)

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    data = tracts_ll,
    fill = FALSE,
    color = "black",
    weight = 1,
    group = "Census tracts"
  ) |>
  addPolygons(
    data = tracts_buf50_ll,
    fill = FALSE,
    color = "red",
    weight = 1,
    group = "Tract buffer 50 m"
  ) |>
  addPolylines(
    data = cycle_ll,
    color = "blue",
    weight = 2,
    group = "Bike lanes"
  ) |>
  addPolygons(
    data = cycle_buf200_ll,
    fillColor = "blue",
    fillOpacity = 0.15,
    color = "blue",
    weight = 1,
    group = "Bike lane buffer 200 m"
  ) |>
  addLayersControl(
    overlayGroups = c(
      "Census tracts",
      "Tract buffer 50 m",
      "Bike lanes",
      "Bike lane buffer 200 m"
    ),
    options = layersControlOptions(collapsed = FALSE)
  )
```

```{r}
library(sf)
library(igraph)
library(dplyr)
library(leaflet)

# prepare segments
segments <- cycle |>
  st_transform(25831) |>
  st_cast("LINESTRING")

# 5 m buffer on each side = connect segments up to 10 m apart
buf <- st_buffer(segments, 15)

# identify which buffered segments touch
touch <- st_intersects(buf)

# build graph of segment connections
g <- graph_from_adj_list(touch, mode = "all")

# component id for each segment
segments$component <- components(g)$membership

# segment length
segments$len <- as.numeric(st_length(segments))

# largest connected component by total length
lcc_id <- segments |>
  st_drop_geometry() |>
  summarise(len = sum(len), .by = component) |>
  slice_max(len, n = 1) |>
  pull(component)

# mark LCC
segments$LCC <- segments$component == lcc_id

# connectivity index
connectivity <- sum(segments$len[segments$LCC]) / sum(segments$len)

# map
segments_ll <- st_transform(segments, 4326)

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolylines(data = segments_ll, color = "blue", weight = 1) |>
  addPolylines(data = segments_ll[segments_ll$LCC, ], color = "red", weight = 3)
```

```{r}
# prepare census tracts
tracts_m <- tracts |>
  st_transform(25831) |>
  select(MUNDISSEC)

# split segments by tract
segments_tract <- st_intersection(
  segments,
  tracts_m
)

# recalculate length after splitting
segments_tract$len <- as.numeric(st_length(segments_tract))

# compute connectivity per tract
connectivity_tract <- segments_tract |>
  st_drop_geometry() |>
  summarise(
    total_len = sum(len),
    lcc_len = sum(len[LCC]),
    .by = MUNDISSEC
  ) |>
  mutate(
    connectivity = lcc_len / total_len)

# join back to tract geometry
tracts_connectivity <- tracts_m |>
  left_join(connectivity_tract, by = "MUNDISSEC")

# transform for leaflet
tracts_ll <- st_transform(tracts_connectivity, 4326)

# map
pal <- colorNumeric("YlOrRd", domain = tracts_ll$connectivity, na.color = "transparent")

leaflet(tracts_ll) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor = ~pal(connectivity),
    fillOpacity = 0.7,
    color = "grey40",
    weight = 0.5,
    popup = ~paste0(
      "MUNDISSEC: ", MUNDISSEC,
      "<br>Connectivity: ", round(connectivity, 3)
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~connectivity,
    position = "bottomright",
    title = "Connectivity",
    opacity = 0.7
  )
```

```{r}
library(sf)
library(igraph)
library(dplyr)

# census tracts
tracts <- st_read(
  "data/Revisió_connectivitat.gpkg",
  layer = "Seccions_censals",
  quiet = TRUE
)

tracts_m <- tracts |>
  st_transform(25831) |>
  select(MUNDISSEC)

years <- 2017:2025

results_tract <- vector("list", length(years))
results_city  <- vector("list", length(years))

for (i in seq_along(years)) {

  cycle_year <- st_read(
    paste0("data/", years[i], "_4T_CARRIL_BICI/", years[i], "_4T_CARRIL_BICI.shp"),
    quiet = TRUE
  )

  segments <- cycle_year |>
    st_transform(25831) |>
    st_cast("LINESTRING")

  buf <- st_buffer(segments, 15)

  touch <- st_intersects(buf)
  g <- graph_from_adj_list(touch, mode = "all")

  segments$component <- components(g)$membership
  segments$len <- as.numeric(st_length(segments))

  lcc_id <- segments |>
    st_drop_geometry() |>
    summarise(len = sum(len), .by = component) |>
    slice_max(len, n = 1) |>
    pull(component)

  segments$LCC <- segments$component == lcc_id

  # city-level connectivity
  results_city[[i]] <- segments |>
    st_drop_geometry() |>
    summarise(
      total_len = sum(len),
      lcc_len = sum(len[LCC])
    ) |>
    mutate(
      connectivity = lcc_len / total_len,
      year = years[i]
    )

  # tract-level connectivity
  segments_tract <- st_intersection(segments, tracts_m)
  segments_tract$len <- as.numeric(st_length(segments_tract))

  results_tract[[i]] <- segments_tract |>
    st_drop_geometry() |>
    summarise(
      total_len = sum(len),
      lcc_len = sum(len[LCC]),
      .by = MUNDISSEC
    ) |>
    mutate(
      connectivity = lcc_len / total_len,
      year = years[i]
    )
}

connectivity_all_years <- bind_rows(results_tract)
connectivity_city <- bind_rows(results_city)
```

